---
title: "Course Project: Weight Lifting Exercise Prediction"
author: "(Haider Ali)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
set.seed(123)
```

# 1. Goal (compact)

Predict the manner in which exercises were performed (`classe` variable) using accelerometer data from multiple sensors. Deliverables: model, short writeup (<2000 words), <5 figures, predictions for 20 test cases (one file per prediction for the Quiz portion).

# 2. Data fetching and quick inspection

```{r data-load}
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
test_url  <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
train_file <- "pml-training.csv"
test_file  <- "pml-testing.csv"

if(!file.exists(train_file)) download.file(train_url, train_file, quiet = TRUE)
if(!file.exists(test_file))  download.file(test_url, test_file, quiet = TRUE)

train <- read.csv(train_file, na.strings = c("", "NA"))
test  <- read.csv(test_file, na.strings = c("", "NA"))

cat('Training rows:', nrow(train), 'columns:', ncol(train), '\n')
cat('Test rows:', nrow(test), 'columns:', ncol(test), '\n')
```

# 3. Preprocessing / feature selection

Short strategy (implemented below):

- Remove near-zero variance predictors.
- Remove predictors with >95% missing values.
- Remove ID / timestamp / user identification columns that don't generalize.
- Optionally scale/center (not strictly necessary for random forest).

```{r preprocess}
library(caret)

# 1) Remove near-zero variance
nzv <- nearZeroVar(train, saveMetrics = TRUE)

# 2) Remove columns with many NAs (>95%)
na_pct <- sapply(train, function(x) mean(is.na(x)))
keep_cols <- names(na_pct[na_pct < 0.95])

# 3) Remove identifiers and non-predictive columns
id_cols <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp", "new_window", "num_window")
keep_cols <- setdiff(keep_cols, id_cols)

# Keep the outcome + selected predictors
keep_cols <- intersect(c(keep_cols, "classe"), names(train))
train2 <- train[, keep_cols]

# Remove any remaining near-zero variance predictors on reduced set
nzv2 <- nearZeroVar(train2)
if(length(nzv2) > 0) train2 <- train2[, -nzv2]

# For caret, ensure factor outcome
train2$classe <- as.factor(train2$classe)

# Apply same column selection to test (careful: test doesn't have classe)
common_cols <- intersect(names(train2), names(test))
test2 <- test[, common_cols]

# Some models require no NAs: we'll impute simple median for numeric predictors
preProcNA <- preProcess(train2[, setdiff(names(train2), "classe")], method = c('medianImpute'))
train_pred <- predict(preProcNA, train2[, setdiff(names(train2), "classe")])
train_final <- data.frame(train_pred, classe = train2$classe)

# Impute test
if(ncol(test2) > 0) {
  test_pred <- predict(preProcNA, test2)
} else {
  test_pred <- test2
}

cat('Predictor columns used:', ncol(train_pred), '\n')
```

# 4. Create an internal train/validation split (for estimating out-of-sample error)

```{r split}
set.seed(123)
inTrain <- createDataPartition(train_final$classe, p = 0.75, list = FALSE)
modelTrain <- train_final[inTrain, ]
modelVal   <- train_final[-inTrain, ]

cat('Model train rows:', nrow(modelTrain), 'Validation rows:', nrow(modelVal), '\n')
```

# 5. Model selection: Random Forest (primary) and GBM (comparison)

Rationale: Random forest handles many predictors, is robust to correlated features, gives variable importance, and usually performs well out-of-the-box for classification on this dataset. We'll still try gradient boosting (gbm) to compare.

```{r models}
ctrl <- trainControl(method = "repeatedcv", number = 5, repeats = 3, classProbs = TRUE)

# Random Forest
set.seed(123)
rfFit <- train(classe ~ ., data = modelTrain, method = "rf", trControl = ctrl, importance = TRUE, ntree = 500)
print(rfFit)

# GBM (comparison)
set.seed(123)
gbmFit <- train(classe ~ ., data = modelTrain, method = "gbm", trControl = ctrl, verbose = FALSE)
print(gbmFit)

# Validation performance
rfPred <- predict(rfFit, modelVal)
gbmPred <- predict(gbmFit, modelVal)

rfCM <- confusionMatrix(rfPred, modelVal$classe)
gbmCM <- confusionMatrix(gbmPred, modelVal$classe)

cat('RF validation accuracy:', rfCM$overall['Accuracy'], '\n')
cat('GBM validation accuracy:', gbmCM$overall['Accuracy'], '\n')
```

# 6. Choose final model and train on full training set

Based on validation accuracy (see above) choose the better model. For reproducibility we'll use Random Forest if its accuracy is higher (typical for this dataset).

```{r final-fit}
# Train final model on all available training data
set.seed(123)
finalFit <- train(classe ~ ., data = train_final, method = "rf", trControl = ctrl, ntree = 500)
finalFit

# Variable importance
varImpRF <- varImp(finalFit)
plot(varImpRF, top = 20)
```

# 7. Expected out-of-sample error

Short write-up (within 2000 words):

- We used repeated 5-fold cross-validation during model tuning to estimate generalization performance and reduce variance in the estimate. The internal hold-out validation accuracy (above) gives a realistic estimate of expected out-of-sample accuracy. In practice, random forest's OOB error and cross-validated accuracy are both informative. Use the validation accuracy reported in the results; the expected out-of-sample error is approximately `r round(1 - rfCM$overall['Accuracy'], 3)` on the held-out portion of the original training set.

# 8. Predictions on provided test set (20 cases)

```{r predict-test}
# Ensure test_pred has same columns as training predictors
missing_cols <- setdiff(names(train_pred), names(test_pred))
if(length(missing_cols) > 0) {
  # Add missing columns with zeros (or NA-imputed values)
  for(col in missing_cols) test_pred[[col]] <- 0
}
# Order columns same as train_pred
test_pred <- test_pred[, names(train_pred)]

# Predict
testPred <- predict(finalFit, test_pred)

# Show predicted classes
testPred

# For the Course Project Prediction Quiz: write each prediction to a separate text file named "problem_id_i.txt"
write_predictions <- function(predictions, outdir = "predictions"){
  if(!dir.exists(outdir)) dir.create(outdir)
  for(i in seq_along(predictions)){
    fname <- file.path(outdir, paste0("problem_id_", i, ".txt"))
    write.table(predictions[i], file = fname, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}

write_predictions(as.character(testPred))
cat('Wrote', length(testPred), 'prediction files to ./predictions\n')
```

# 9. Reproducibility notes

- `set.seed(123)` set at top for reproducibility.
- Session info printed below for exact package versions.

```{r session-info}
sessionInfo()
```

# 10. Short conclusions (keep concise)

- Random Forest gave strong accuracy on validation; expected out-of-sample error is low (see section 7).
- Important predictors are captured in `varImp` plot; accelerometer-derived summary features are highly informative.
- Limitations: we removed many variables with >95% missing values — more sophisticated imputation or feature engineering might improve performance. Also, model interpretability is limited; consider examining partial dependence or training simpler learners if interpretability is required.

---

*End of analysis.*
